---
title: "Sports Rorts"
author: "Jeremy Forbes"
date: "19/02/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(visreg)
library(mgcv)
load("df.rda")
```

## Sports Rorts - Uncovering Government Bias with Statistical Modelling

You can run, but you can't hide (from statistics). The Australian Liberal party is finding this out the hard way.

In recent weeks, the Liberal party has been accused of using $100M of sporting grants to gain votes in the lead up to the 2019 election. As a data scientist, I believe there's only one way to find out if this is true - by crunching the numbers. And so, I set out to see what can be discovered with publicly available data relating to the "Sports Rorts" scandal.

My thinking is that if the Liberal party allocated grants to electorates on the basis of their voting behaviour, then this strategy could be revealed with an appropriate statistical model.

Now before I go any further, here's some background in case you haven't been reading the news...

#### Background

The Australian Liberal party has been in government since the 2013 federal election. In the lead up to last year's election, the Liberal party announced a community sport infrastructure program, offering grants of up to $500,000 for local sporting organisations. This program was marketed as an initiative to boost sporting participation across the country. Organisations had to submit an application, which would be independently assessed and scored by Sport Australia. These scores were to be used by the Sports Minister in deciding which grants would be approved.

This year an audit on the funding scheme was conducted by the Australian National Audit Office (ANAO). Their report implied (but did not confirm) that the government, led by the Liberal party, gave preference to grant applications from 'marginal' and 'targeted' electorates, rather than following the advice from Sport Australia. Essentially, the Liberal party are accused of adopting a strategic plan to win support in specific electorates (seats) by approving more grants in those areas, in order to order to help them secure victory in the 2019 election.

#### Back to the action

The aim of my analysis is to see whether there is evidence that the Liberal party allocated grants on the basis of the electoral voting behaviour from the previous elecction. Essentially, is there an underlying strategy? And if so, what voting pattern was targeted?

To go ahead with this analysis, I need to combine three key data sources, all of which are publicly available:
(1) Information about each grant that was approved (location and $ amount),
(2) Election results for each electorate from the previous election (2016), and
(3) Census data to control for other socio-demographic factors that might be at play.

With this information, I'll be able to produce a dataset with the total grant money allocated to each electorate, along with its previous voting behaviour and other controls from the Census.

(Note that all steps of data collection, wrangling and analysis are conducted using R.)

#### Obtaining the data

Sports Australia has published a list of the 684 approved grants in a table on their website (https://www.sportaus.gov.au/grants_and_funding/community_sport_infrastructure_grant_program/successful_grant_recipient_list), so I scraped this using *rvest*. Next, this list is to a Google Maps API in order to get the lat-long coordinates of each grant (using *ggmap* from https://github.com/dkahle/ggmap). This works for almost all of the grants, and some quick googling helps find where the remainder of these organisations are located.

To allocate each sporting organisation to its corresponding seat, I need a map of electorates from the 2016 federal election. Luckily, this is readily available in the `eechidna` package, along with the electoral voting data. Using some nifty functions from `sp` and `rgeos` to loop through the electorates, each grant is successfully allocated.

The grants are now aggregated so that the resultant dataset holds the total grant $ amount for each of the 150 federal electorates. This is then joined with the two-party preferred vote from the 2016 election, the swing vote from the 2016 election, and a collection of socio-demographic variables from the 2016 Census - all obtained from `eechidna`. (The two-party preferred vote is a measure of support between only the Liberal and Labor parties, and the swing vote is the percentage point change in the vote from 2013 to 2016 elections for that seat).

Here's a snapshot of the resultant dataset.

```{r}
df %>% 
  select(-Number_Grants) %>% 
  head
```

Now time to get stuck into the analysis.

#### The fun stuff

What I'm really looking to uncover is the relationship between the percentage of votes in favour of the Liberal party for each electorate in the 2016 election, and the amount of grant money allocated to that electorate. First, let's see what a scatterplot looks like with a linear regression line fitted.

```{r}
df %>% 
  ggplot(aes(x = LNP_Percent, y = Amount)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "% of Votes for Coalition", y = "Total Grant Money ($)")
```

There's some evidence here to suggest that electorates with stronger support for the Coalition were more likely to receive sports grant money. But really, its likely that if the Coalition did implement a targeted strategy in allocating grants, the relationship between these two variables would not be linear. A strategy that gives the most money to seats that are the safest wouldn't be a very good one - because its highly unlikely that these seats would be vulnerable to changing hands at the 2019 election.

So instead of fitting a linear regression line, let's plot a really flexible regression curve. For this I have chosen to use the non-parametric loess curve, which is a smooth curve that gives higher weighting to local data points. This allows us to get a better picture of the relationship between the 2016 vote and grant amount, rather than imposing a linear restriction.

```{r}
df %>% 
  ggplot(aes(x = LNP_Percent, y = Amount)) +
  geom_point() +
  geom_smooth(span = 0.5) +
  labs(x = "% of Votes for Coalition", y = "Total Grant Money ($)")
```

Wow! The loess curve is revealing a pretty significant relationship - which was not obvious from inspecting the initial scatterplot. There is a clear spike in the amount of grant money given to electorates that were "marginal" Coalition seats in the 2016 election. "Marginal" Coalition seats are those that were won by the Coalition, but weren't won by a huge margin. I think this plot itself is pretty convincing evidence of a deliberate strategy in allocating the grants.

But to take this another step further, I decided to fit a GAM (Generalized Additive Model) to model the grant amount as a function of the 2016 vote, as well as Census characteristics of the electorates and the swing vote. Each of these variables is included in the model as a smooth term. In this case I have used penalized regression splines - a type of non-parametric regression, similar to loess, that allows for patterns in the data to reveal itself.

By controlling for these other variables, the effect of the vote itself can be identified. It is possible that once these other effects are accounted for, we will no longer see a relationship between the vote and grant amounts. Let's go ahead and see...

```{r}
mod <- gam(Amount ~ s(LNP_Percent) + s(Population) + s(MedianAge) + 
    s(MedianPersonalIncome) + s(HighSchool) + s(Unemployed) + s(Owned) + s(Swing),
  data = df)
visreg(mod, "LNP_Percent", alpha = 0.05, gg = T) +
  labs(x = "% of Votes for Coalition", y = "Fitted Model - Total Grant Money ($)")

mod2 <- gam(Amount ~ s(Population) + s(MedianAge) + 
    s(MedianPersonalIncome) + s(HighSchool) + s(Unemployed) + s(Owned),
  data = df)
```

It can be clearly seen that after accounting for these other variables, there is still a huge spike in grant amount for marginal Coalition seats! Having a look at the p-values from the GAM regression output, there is significant evidence that the 2016 vote influenced the allocation of grant money across electorates (p = 0.002). The resultant model itself explains 46.3% of the deviance - which is a good fit considering that the official aim of the grant scheme was set out to increase sport engagement Australia-wide, and the applications were scored by an independent body. 

I also fit a model omitting the voting data (both two-party preferred and swing vote), and compared the two models with a Chi-square test. The model including the election data was a significantly better fit (p = 0.000005). On this basis, the evidence of a strategic bias is quite compelling.

To check that my findings are robust to outliers, I re-estimated the model after removing the two electorates with outstandingly large grant amounts ($1.8M and $2.0M - the electorates of Boothby and Dawson), and found that the same effect were observed. Additionally, the model assumptions seem to be valid, as the residuals are approximately normal (observed using a quantile-quantile plot).

#### In conclusion

My statistical modelling approach provides empirical evidence that the Liberal party did in fact allocate sports grants on the basis of an underlying strategy to beef up support in the lead up to the 2019 election. In a classic case of pork barreling, marginal seats that were won by the Liberal party received significantly higher grant allocations, in what the model suggests was an attempt to secure these electorates for the 2019 election.

Not only is this a pretty cool finding, but importantly this demonstrates how readily powerful insights can be revealed using publicly available data. 

All of the data collection, wrangling and analysis was conducted using `R`. For anyone interested in exploring Australian election and Census data, I highly recommend checking out the `eechidna` package (of which I am an author and maintainer). This package makes it easy to access Australian election and Census data from 2001-2019 (including maps), available on CRAN.


## REMOVE BELOW


In recent weeks, the Australian Liberal party has been under intense scrutiny for its allocation of grants to sporting organisations under a scheme that took place in the lead up to last year's federal election. Dubbed the "Sports Rorts" affair, the Liberal party (Coalition) is accused of approving grants on the basis of an underlying strategy that aimed to boost support in targeted electorates.

As a data scientist, I believe that this accusation is a question that can be readily explored using publicly available data and the pragmatic application of statistics. My theory is that if electorates were in fact given preference on the basis of their previous voting behaviour, then the Liberal party's strategy would be revealed by fitting an appropriate statistical model.


#### Checking that my findings are robust

Looking at the distribution of grant amounts across electorates with a simple boxplot reveals that there are two outliers with grant amounts of $1.8M and $2.0M - the electorates of Boothby and Dawson. I refit the model after removing those two electorates from the dataset, and still find that the effect of the 2016 vote is significant (p = 0.008), and the spike in grant amount still remains.

```{r}
mod_robust <- gam(Amount ~ s(LNP_Percent) + s(Population) + s(MedianAge) + 
    s(MedianPersonalIncome) + s(HighSchool) + s(Unemployed) + s(Owned) + s(Swing),
  data = df %>% 
    filter(Amount < 1600000))
visreg(mod_robust, "LNP_Percent", alpha = 0.05)
```

Similarly, when removing the seats with high Cook's Distance and refitting the model, the same effect is observed.

There also don't seem to be any material issues with the residuals as they appear approximately normal in a quantile-quantile plot.

```{r}
# Quantile-quantile plot
ggplot(aes(sample = mod$residuals), data = NULL) + 
  geom_qq() + 
  geom_qq_line()
```


Check out my #medium debut on my analysis of the "Sports Rots" scandal - a statistical modelling to see how sporting grant money was allocated on the basis of previous election results, all done using the #eechidna #Rstats package and a Google Maps API.
